```{r set}
#| message: false
#| include: false
library(tidyverse)
library(ggfortify)
```

# Climate {.unnumbered}

We retrieved all data from **TerraClimate** [@abatzoglou2018]. The variables are extracted for each plot of each site and every month from 1958 to 2023. The raw variables are:

-   ET: Actual evapotranspiration, mm
-   CWD: Climate water deficit, mm
-   Tmax: Maximum temperature, °C
-   Tmin: Minimum temperature, °C
-   PDSI: Palmer Drought Severity Index
-   PET: Potential evapotranpiration, mm
-   Pr: precipitation, mm
-   Soil: Soil humidity, mm
-   VPD: Vapour Pressure Deficit, kPa

```{r prepclim}
#| eval: false
all_dat <- list.files("data/derived_data/climate/", full.names = TRUE) %>%
  read_tsv() %>%
  group_by(site) %>%
  mutate(
    aet = aet * .1, def = def * .1, pdsi = pdsi * .01, pet = pet * .1,
    tmmn = tmmn * .1, tmmx = tmmx * .1, vpd = vpd * .01, soil = soil * .1
  ) %>%
  mutate(month = month(time)) %>%
  mutate(year = year(time))
monthly <- all_dat %>%
  select(site, month, pr, tmmn, tmmx, soil) %>%
  group_by(site, month) %>%
  summarise_all(mean)
write_tsv(monthly, "data/derived_data/climate_month.tsv")
ds <- all_dat %>%
  filter(aet > pr) %>%
  group_by(site, year) %>%
  summarise(dsl = n() * 30, dsi = sum(aet - pr)) %>%
  select(-year) %>%
  group_by(site) %>%
  summarise_all(mean)
yearly <- all_dat %>%
  group_by(site, year) %>%
  summarise(
    et = sum(aet),
    cwd = sum(def),
    pr = sum(pr),
    tmax = max(tmmx),
    vpd = max(vpd),
    soil = min(soil)
  ) %>%
  select(-year) %>%
  group_by(site) %>%
  summarise_all(mean) %>%
  left_join(ds)
write_tsv(yearly, "data/derived_data/climate_year.tsv")
```

For a quick exploration of each site climatology we first had a look to umbrothermal diagrams for the mean climate of each site, opposing for instance highly seasonal monsoon regimes from India un Uppangala to less seasonal regimes from Misiones in Argentina.

```{r umbrothermal}
#| message: false
#| warning: false
#| fig-cap: "Umbrothermal diagrams for the mean climate of each site."
read_tsv("data/derived_data/climate_month.tsv") %>%
  filter(!(site %in% c("Uppangala", "Nelliyampathy"))) %>%
  group_by(site, month) %>%
  summarise_all(mean) %>%
  ggplot(aes(x = month)) +
  geom_col(aes(y = pr / 10, fill = pr < 100), col = NA) +
  geom_line(aes(y = tmmn, col = "maximum"), linewidth = 1.1) +
  geom_line(aes(y = tmmx, col = "minimum"), linewidth = 1.1) +
  theme_bw() +
  xlab("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")
  ) +
  scale_y_continuous(
    name = "Temperature (°C)",
    sec.axis = sec_axis(trans = ~ . * 10, name = "Precipitation (mm)")
  ) +
  scale_color_discrete(guide = "none") +
  facet_wrap(~site) +
  scale_fill_manual(guide = "none", values = c("darkgrey", "lightgrey"))
```

```{r umbrothermal_uppangala}
#| message: false
#| warning: false
#| fig-cap: "Umbrothermal diagrams for the mean climate of each site."
read_tsv("data/derived_data/climate_month.tsv") %>%
  filter(site %in% c("Uppangala", "Nelliyampathy")) %>%
  group_by(site, month) %>%
  summarise_all(mean) %>%
  ggplot(aes(x = month)) +
  geom_col(aes(y = pr / 10, fill = pr < 100), col = NA) +
  geom_line(aes(y = tmmn, col = "maximum"), linewidth = 1.1) +
  geom_line(aes(y = tmmx, col = "minimum"), linewidth = 1.1) +
  theme_bw() +
  xlab("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")
  ) +
  scale_y_continuous(
    name = "Temperature (°C)",
    sec.axis = sec_axis(trans = ~ . * 10, name = "Precipitation (mm)")
  ) +
  scale_color_discrete(guide = "none") +
  facet_wrap(~site) +
  scale_fill_manual(guide = "none", values = c("darkgrey", "lightgrey"))
```

To ease plot and sites comparisons, we derived yearly metrics:

-   CWD: Climate water deficit, mm, mean across years of the sum of monthly CWD
-   ET: Evapotranspiration, mm, mean across years of the sum of monthly ET
-   Pr: Precipitation, mm, mean across years of the sum of monthly Pr
-   Soil: Soil humidity, mm, mean across years of the minimum of monthly soil humidity
-   VPD: Vapour Pressure Deficit, kPa, mean across years of the maximum of monthly VPD
-   Tmax: Maximum temperature, °C, mean across years of the maximum of monthly Tmax
-   DSL: Dry season length, days, mean across years of the number of month with ET\>Pr multiplied by 30
-   DSI: Dry season intensity, mm, mean across years of the sum of ET-Pr for month with ET\>Pr

This better revealed among sites differences with for instance highest dry season intensity and length at Uppangala despite highest total precipitation.

```{r metrics}
#| message: false
#| warning: false
#| fig-cap: "Yearly mean climate variables per site."
read_tsv("data/derived_data/climate_year.tsv") %>%
  gather(variable, value, -site) %>%
  mutate(var_long = recode(variable,
    "et" = "Evapotranspiration [ mm ]",
    "cwd" = "Climate water deficit [ mm ]",
    "pr" = "Precipitation [ mm ]",
    "tmax" = "Max temperature [ °C ]",
    "vpd" = "Max VPD [ kPA ]",
    "soil" = "Min soil humidity [ mm ]",
    "dsi" = "Dry season intensity [ mm ]",
    "dsl" = "Dry season length [ day ]"
  )) %>%
  ggplot(aes(site, value, fill = site, group = site)) +
  geom_col(position = "dodge") +
  facet_wrap(~var_long, scales = "free_y") +
  theme_bw() +
  theme(
    legend.position = "bottom", axis.title = element_blank(),
    axis.text.x = element_blank(),
    legend.key.size = unit(0.5, "line")
  ) +
  scale_fill_discrete("")
```

The first ordination axis retained almost half of the variations.

```{r pca}
#| message: false
#| warning: false
#| fig-cap: "Yearly mean climate variables PCA per site and plot."
data <- read_tsv("data/derived_data/climate_year.tsv")
pca <- prcomp(data %>% select(-site), scale. = TRUE)
autoplot(pca,
  loadings = TRUE, loadings.label = TRUE,
  loadings.label.repel = TRUE,
  data = data, colour = "site"
) +
  theme_bw() +
  coord_equal() +
  scale_color_discrete("") +
  theme(legend.key.size = unit(0.5, "line"))
```

```{r corr}
#| message: false
#| warning: false
#| fig-cap: "Climate variables pairwise correlations."
read_tsv("data/derived_data/climate_year.tsv") %>%
  select(-site) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(type = "upper", diag = FALSE)
```

We will extract all for modelling but group 2 is focusing on climate stress which I think could be well represented by DSL, DSI, Soil and/or Tmax.
