```{r setup}
#| message: false
#| include: false
library(tidyverse)
```

# Soil {.unnumbered}

We retrieved all data from SoilGrids detailed below [@poggio2021], but the spatial extrapolation and reference points have some issues as for instance in French Guiana. This could be compared to site data, but with issues regarding site data standardization, representativeness, and validation.

-   Clay: Proportion of clay particles (\< 0.002 mm) in the fine earth fraction, %
-   Sand: Proportion of sand particles (\> 0.05 mm) in the fine earth fraction, %
-   Silt: Proportion of silt particles (≥ 0.002 mm and ≤ 0.05 mm) in the fine earth fraction, %
-   BD: Bulk density of the fine earth fraction, kg dm³
-   CEC: Cation Exchange Capacity of the soil, cmolC kg^-1^
-   CF: Volumetric fraction of coarse fragments (\> 2 mm), cm^3^ 100cm^-3^
-   N: Total nitrogen, g kg^-1^
-   pH: Soil pH
-   SOC: Soil organic carbon content in the fine earth fraction, g kg^-1^
-   OCD: Organic carbon density, kg dm³
-   OCS: Organic carbon stocks, kg m^2^

```{r prep}
#| eval: false
list.files("data/soil/", full.names = TRUE) %>%
  read_tsv() %>%
  select(-time) %>%
  gather(variable, value, -site, -plot, -lon, -lat) %>%
  separate(variable, c("variable", "depth"), "_") %>%
  separate(depth, c("mindepth", "depth"), "-") %>%
  mutate(depth = as.numeric(gsub("cm", "", depth))) %>%
  select(-mindepth) %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  mutate(
    bdod = bdod / 100,
    cec = cec / 10,
    cfvo = cfvo / 10,
    clay = clay / 10,
    sand = sand / 10,
    silt = silt / 10,
    nitrogen = nitrogen / 100,
    phh2o = phh2o / 10,
    soc = soc / 10,
    ocd = ocd / 10,
    ocs = ocs / 10
  ) %>%
  write_tsv("outputs/soil.tsv")
```

> BAFOG and Paracou are lacking because too close to the sea.

The plots showed low variation within site but wider variation among sites with for instance sandy soils from Iwokrama opposed to silt and clay soils from Montagne Tortue.

```{r allplot}
#| message: false
#| warning: false
#| fig-cap: "Mean soil data per site and depth."
read_tsv("outputs/soil.tsv") %>%
  select(-lon, -lat, -plot) %>%
  group_by(site, depth) %>%
  summarise_all(mean, na.omit = TRUE) %>%
  gather(variable, value, -site, -depth) %>%
  mutate(var_long = recode(variable,
    "bdod" = '"BD ["~kg~dm^{-3}~"]"',
    "cec" = '"CEC ["~cmol~kg^{-1}~"]"',
    "cfvo" = '"CF ["~cm^3~"100"~cm^{-3}~"]"',
    "clay" = '"Clay ["~"%"~"]"',
    "nitrogen" = '"N ["~g~kg^{-1}~"]"',
    "ocd" = '"OCD ["~kg~dm^{3}~"]"',
    "ocs" = '"OCS ["~kg~m^{-2}~"]"',
    "phh2o" = "pH",
    "sand" = '"Sand ["~"%"~"]"',
    "silt" = '"Silt ["~"%"~"]"',
    "soc" = '"SOC ["~g~kg^{-1}~"]"'
  )) %>%
  ggplot(aes(site, value, fill = depth)) +
  geom_col(aes(group = as.factor(depth)), position = "dodge") +
  facet_wrap(~var_long, scales = "free_x", labeller = label_parsed) +
  theme_bw() +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_text(size = 5)
  ) +
  scale_fill_viridis_c() +
  coord_flip()
```

```{r allplot5}
#| message: false
#| warning: false
#| fig-cap: "Mean soil data per site for top horizon (up to 5cm)."
read_tsv("outputs/soil.tsv") %>%
  select(-lon, -lat, -plot) %>%
  group_by(site, depth) %>%
  summarise_all(mean, na.omit = TRUE) %>%
  filter(depth == 5) %>%
  gather(variable, value, -site, -depth) %>%
  na.omit() %>%
  mutate(var_long = recode(variable,
    "bdod" = '"BD ["~kg~dm^{-3}~"]"',
    "cec" = '"CEC ["~cmol~kg^{-1}~"]"',
    "cfvo" = '"CF ["~cm^3~"100"~cm^{-3}~"]"',
    "clay" = '"Clay ["~"%"~"]"',
    "nitrogen" = '"N ["~g~kg^{-1}~"]"',
    "ocd" = '"OCD ["~kg~dm^{3}~"]"',
    "ocs" = '"OCS ["~kg~m^{-2}~"]"',
    "phh2o" = "pH",
    "sand" = '"Sand ["~"%"~"]"',
    "silt" = '"Silt ["~"%"~"]"',
    "soc" = '"SOC ["~g~kg^{-1}~"]"'
  )) %>%
  ggplot(aes(site, value, fill = site)) +
  geom_col(aes(group = as.factor(depth)), position = "dodge") +
  facet_wrap(~var_long, scales = "free_y", labeller = label_parsed) +
  theme_bw() +
  theme(
    legend.position = "bottom", axis.title = element_blank(),
    axis.text.x = element_blank()
  ) +
  scale_fill_discrete("")
```

```{r betterplot}
#| message: false
#| warning: false
#| fig-cap: "Soil texture for all sites, plots and depths."
read_tsv("outputs/soil.tsv") %>%
  ggplot(aes(x = sand, y = clay, z = silt, col = site)) +
  ggtern::coord_tern(L = "x", T = "y", R = "z") +
  geom_point() +
  theme_bw() +
  labs(
    yarrow = "Clay [ % ]",
    zarrow = "Silt [ % ]",
    xarrow = "Sand [ % ]"
  ) +
  ggtern::theme_showarrows() +
  ggtern::theme_hidetitles() +
  ggtern::theme_clockwise() +
  scale_color_discrete("") +
  theme(legend.position = "bottom")
```

The first ordination axis retained almost half of the variations of the 10 soil variables and opposed sandy poor soils to more fertile silty soils.

> This is a very quick interpretation of the figure and should be explored more in depth.

```{r pca}
#| message: false
#| warning: false
#| fig-cap: "Soil variables PCA per site and plot."
library(ggfortify)
data <- read_tsv("outputs/soil.tsv") %>%
  filter(depth == 5) %>%
  select(-ocs) %>%
  na.omit()
pca <- prcomp(data %>% select(-site, -plot, -lat, -lon, -depth), scale. = TRUE)
autoplot(pca,
  loadings = TRUE, loadings.label = TRUE,
  loadings.label.repel = TRUE,
  data = data, colour = "site"
) +
  theme_bw() +
  coord_equal() +
  scale_color_discrete("")
```
