```{r set}
#| message: false
#| include: false
library(tidyverse)
```

# Landscape {.unnumbered}

We used information on total forest cover area in a radius of 1-km from **TMF** [@europeancommission.jointresearchcentre.2020].

```{r prep}
#| eval: false
list.files("data/derived_data/landscape/", full.names = TRUE) %>%
  read_tsv() %>%
  group_by(site, plot, year, forest) %>%
  summarise(n = n()) %>%
  group_by(site, plot, year) %>%
  mutate(n_tot = sum(n)) %>%
  pivot_wider(names_from = forest, values_from = n) %>%
  select(-`0`) %>%
  mutate(`1` = ifelse(is.na(`1`), 0, `1`)) %>%
  summarise(intact = `1` / n_tot * 100) %>%
  mutate(intact = ifelse(site == "Misiones", NA, intact)) %>%
  mutate(intact = ifelse(site == "Kibale", NA, intact)) %>%
  write_tsv("data/derived_data/landscape.tsv")
```

Most sites showed low deforestation level in the 90s but most of them showed deforestation increasing through time.

> There are no data for Misiones as it is considered not in TMF by JRC and no data for Kibale because the coordinates are not precise enough.

```{r allsites}
#| message: false
#| warning: false
#| fig-cap: "Forest cover through time."
read_tsv("data/derived_data/landscape.tsv") %>%
  na.omit() %>%
  ggplot(aes(year, 100 - intact, col = site, group = paste(site, plot))) +
  geom_line() +
  theme_bw() +
  xlab("") +
  ylab("Deforested area in a radius of 1-km [ % ]") +
  facet_wrap(~site) +
  scale_color_discrete(guide = "none") +
  scale_y_sqrt()
```

Most sites showed less than 10% forest loss cover in 2022, but few sites such as Tapajos km67 lost more than a third of their forest cover.

```{r 2022cover}
#| message: false
#| warning: false
#| fig-cap: "2022 forest cover."
read_tsv("data/derived_data/landscape.tsv") %>%
  filter(year == 2022) %>%
  ggplot(aes(site, 100 - intact, fill = site, group = paste(site, plot))) +
  geom_col(position = "dodge") +
  theme_bw() +
  xlab("") +
  ylab("Deforested area in 2022 in a radius of 1-km [ % ]") +
  scale_fill_discrete(" ") +
  coord_flip() +
  scale_y_sqrt() +
  theme(legend.key.size = unit(0.5, "line"))
```

Almost all sites lost forest cover between 1990 and 2022 at the exception of few plots in Sao Nicolau that gained almost 20% of cover.

```{r coverchange}
#| message: false
#| warning: false
#| fig-cap: "Forest cover change between 1990 and 2022."
read_tsv("data/derived_data/landscape.tsv") %>%
  filter(year %in% c(1990, 2022)) %>%
  arrange(site, plot, year) %>%
  group_by(site, plot) %>%
  mutate(delta = intact - lag(intact)) %>%
  na.omit() %>%
  ggplot(aes(site, delta, fill = site, group = paste(site, plot))) +
  geom_col(position = "dodge") +
  theme_bw() +
  xlab("") +
  ylab("Forest surface change between 1990 and
       2022 in a radius of 1-km [ % ]") +
  scale_fill_discrete(" ") +
  coord_flip() +
  theme(legend.key.size = unit(0.5, "line"))
```
