```{r set}
#| message: false
#| include: false
library(tidyverse)
library(ggfortify)
```

# Metrics {.unnumbered}

Metrics gathering for modelling steps of the project:

-   CWD: Climate water deficit, mm
-   ET: Evapotranspiration, mm
-   Pr: Precipitation, mm
-   Soil: Soil humidity, mm
-   VPD: Vapour Pressure Deficit, kPa
-   Tmax: Maximum temperature, °C
-   DSL: Dry season length, daysmultiplied by 30
-   DSI: Dry season intensity, mm
-   Clay: Proportion of clay, %
-   Sand: Proportion of sand, %
-   Silt: Proportion of silt, %
-   BDOD: Bulk density, kg dm³
-   CEC: Cation Exchange Capacity, cmolC kg^-1^
-   CF: Coarse Fragments , cm^3^ 100cm^-3^
-   N: Soil nitrogen, g kg^-1^
-   pH: Soil pH
-   SOC: Soil organic carbon content, g kg^-1^
-   OCD: Organic carbon density, kg dm³

```{r prep}
#| eval: false
sites <- read_tsv("data/derived_data/sites.tsv") %>%
  select(-plot, -latitude, -longitude) %>%
  unique()
climate <- read_tsv("data/derived_data/climate_year.tsv")
soil <- read_tsv("data/derived_data/soil.tsv") %>%
  group_by(site) %>%
  filter(depth <= 15) %>%
  summarise_all(mean, na.omit = TRUE) %>%
  select(-X, -Y, -depth, -ocs)
sites %>%
  left_join(climate) %>%
  left_join(soil) %>%
  write_tsv("outputs/environment.tsv")
```

We have correlation among group of variables, notably among climate and soil, which is logical but should be taken into account. For instance SOC is highly anti-correlated to DSL and DSI.

```{r pca}
#| message: false
#| warning: false
#| fig-cap: "All variables PCA per site and plot."
data <- read_tsv("outputs/environment.tsv") %>%
  na.omit()
pca <- prcomp(data %>% select(-site),
  scale. = TRUE
)
autoplot(pca,
  loadings = TRUE, loadings.label = TRUE,
  loadings.label.repel = TRUE,
  data = data, colour = "site"
) +
  theme_bw() +
  coord_equal() +
  scale_color_discrete("") +
  theme(legend.key.size = unit(0.5, "line"))
```

```{r corr}
#| message: false
#| warning: false
#| fig-cap: "All variables pairwise correlations."
read_tsv("outputs/environment.tsv") %>%
  select(-site) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(type = "upper", diag = FALSE)
```
