---
title: "Site environment"
author: "Schmitt, S."
date: last-modified
format:
  pdf:
    number-sections: false
    link-citations: true
    colorlinks: true
    fig-pos: H
execute:
  echo: false
params:
  site: "Misiones"
bibliography: ../references.bib
---

The purpose of this document is to provide site PIs with the collected environmental, climate and soil data for their site. **You are invited to provide feedback if any climate or soil values seem incorrect to you, or if you have local comparative data**. Bear in mind, however, that this is a medium scale, especially for soil, which is known to vary at a very fine scale.

```{r set}
#| include: false
library(tidyverse)
library(sf)
library(knitr)
options(knitr.kable.NA = "")
```

## Site

Maps below are here to check the given coordinates.

```{r site_maps}
#| message: false
#| warning: false
#| fig-cap: "Site localisation in world and in the country, as well as plots coordinates in the site." #nolint
site <- read_tsv("../data/derived_data/sites.tsv") %>%
  filter(site == params$site) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
sf_use_s2(FALSE)
world <- rnaturalearth::ne_countries(scale = 10, returnclass = "sf")
country <- st_intersection(world, site)$admin %>% unique()
g_world <- ggplot() +
  geom_polygon(
    data = map_data("world"),
    aes(x = long, y = lat, group = group), fill = "lightgrey"
  ) +
  geom_sf(data = site, col = "red", fill = "lightgrey") +
  ggtitle(params$site, country) +
  theme_void()
g_country <- world %>%
  filter(geounit == country) %>%
  ggplot() +
  geom_sf(fill = "lightgrey") +
  geom_sf(data = site, col = "red") +
  theme_void()
g_local <- ggplot() +
  basemapR::base_map(st_bbox(site), basemap = "positron") +
  geom_sf(data = site) +
  ggsflabel::geom_sf_text_repel(data = site, aes(label = plot)) +
  theme_bw() +
  xlab("") +
  ylab("") +
  theme(
    axis.text = element_text(size = 4),
    axis.text.x = element_text(size = 4, angle = 90)
  )
gridExtra::grid.arrange(
  g_world, g_country, g_local,
  widths = c(1, 1.5),
  layout_matrix = rbind(
    c(1, 3),
    c(2, 3)
  )
)
```

## Climate

> [TerraClimate](https://www.climatologylab.org/terraclimate.html) is a dataset of monthly climate and climatic water balance for global terrestrial surfaces from 1958-2019. These data provide important inputs for ecological and hydrological studies at global scales that require high spatial resolution and time-varying data. All data have monthly temporal resolution and a \~4-km (1/24th degree) spatial resolution. The data cover the period from 1958-2020. We plan to update these data periodically (annually). [@abatzoglou2018]

We retrieved all available data from TerraClimate [@abatzoglou2018] of every month from 1958 to 2023. We obtained the following climatology represented through an umbrothermal diagram:

```{r umbrothermal}
#| message: false
#| warning: false
#| fig-cap: "Umbrothermal diagram for the mean climate of the site. Vertical grey bars represents mean monthly precipitation in mm with lightgrey bars representing dry season with less than 100 mm and dark grey bars the rainy season. Lines represents maximal in red and minimal in blue monthly temperature in °C." #nolint
read_tsv("../data/derived_data/climate_month.tsv") %>%
  filter(site == params$site) %>%
  group_by(site, month) %>%
  summarise_all(mean) %>%
  ggplot(aes(x = month)) +
  geom_col(aes(y = pr / 10, fill = pr < 100), col = NA) +
  geom_line(aes(y = tmmn, col = "maximum"), linewidth = 1.1) +
  geom_line(aes(y = tmmx, col = "minimum"), linewidth = 1.1) +
  theme_bw() +
  xlab("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")
  ) +
  scale_y_continuous(
    name = "Temperature (°C)",
    sec.axis = sec_axis(trans = ~ . * 10, name = "Precipitation (mm)")
  ) +
  scale_color_manual(guide = "none", values = c("blue", "red")) +
  facet_wrap(~site) +
  scale_fill_manual(guide = "none", values = c("darkgrey", "lightgrey"))
```

We derived the following yearly metrics:

-   **CWD**: Climate water deficit, mm, mean across years of the sum of monthly CWD
-   **ET**: Evapotranspiration, mm, mean across years of the sum of monthly ET
-   **Pr**: Precipitation, mm, mean across years of the sum of monthly Pr
-   **Soil**: Soil humidity, mm, mean across years of the minimum of monthly soil humidity
-   **VPD**: Vapour Pressure Deficit, kPa, mean across years of the maximum of monthly VPD
-   **Tmax**: Maximum temperature, °C, mean across years of the maximum of monthly Tmax
-   **DSL**: Dry season length, days, mean across years of the number of month with ET\>Pr multiplied by 30
-   **DSI**: Dry season intensity, mm, mean across years of the sum of ET-Pr for month with ET\>Pr

```{r metrics}
#| message: false
#| warning: false
read_tsv("../data/derived_data/climate_year.tsv") %>%
  filter(site == params$site) %>%
  gather(variable, value, -site) %>%
  mutate(acronym = recode(variable,
    "et" = "ET",
    "cwd" = "CWD",
    "pr" = "Pr",
    "tmax" = "Tmax",
    "vpd" = "VPD",
    "soil" = "Soil",
    "dsi" = "DSI",
    "dsl" = "DSL"
  )) %>%
  mutate(unit = recode(variable,
    "et" = "mm",
    "cwd" = "mm",
    "pr" = "mm",
    "tmax" = "°C",
    "vpd" = "kPA",
    "soil" = "mm",
    "dsi" = "mm",
    "dsl" = "day"
  )) %>%
  mutate(variable = recode(variable,
    "et" = "Evapotranspiration",
    "cwd" = "Climate water deficit",
    "pr" = "Precipitation",
    "tmax" = "Maximum temperature",
    "vpd" = "Maximum VPD",
    "soil" = "Minimum soil humidity",
    "dsi" = "Dry season intensity",
    "dsl" = "Dry season length"
  )) %>%
  select(acronym, variable, value, unit) %>%
  arrange(acronym) %>%
  kable(caption = "Yearly mean climate variables for the site.", digits = 2)
```

## Soil

> [SoilGrids](https://www.isric.org/explore/soilgrids) is a system for global digital soil mapping that uses state-of-the-art machine learning methods to map the spatial distribution of soil properties across the globe. SoilGrids prediction models are fitted using over 230 000 soil profile observations from the [WoSIS database](https://www.isric.org/explore/wosis/faq-wosis "Go to WoSIS, FAQ pages") and a series of environmental covariates. Covariates were selected from a pool of over 400 environmental layers from Earth observation derived products and other environmental information including climate, land cover and terrain morphology. The outputs of SoilGrids are global soil property maps at six standard depth intervals (according to the GlobalSoilMap IUSS working group and its specifications) at a spatial resolution of 250 meters. Prediction uncertainty is quantified by the lower and upper limits of a 90% prediction interval. The additional uncertainty layer displayed at [soilgrids.org](https://soilgrids.org/) is the ratio between the inter-quantile range and the median. The SoilGrids maps are publicly available under the [CC-BY 4.0 License](https://creativecommons.org/licenses/by/4.0/). [@poggio2021]

We retrieved all data from **SoilGrids** detailed below [@poggio2021]:

-   **Clay**: Proportion of clay particles (\< 0.002 mm) in the fine earth fraction, %
-   **Sand**: Proportion of sand particles (\> 0.05 mm) in the fine earth fraction, %
-   **Silt**: Proportion of silt particles (≥ 0.002 mm and ≤ 0.05 mm) in the fine earth fraction, %
-   **BD**: Bulk density of the fine earth fraction, kg dm³
-   **CEC**: Cation Exchange Capacity of the soil, cmolC kg^-1^
-   **CF**: Volumetric fraction of coarse fragments (\> 2 mm), cm^3^ 100cm^-3^
-   **N**: Total nitrogen, g kg^-1^
-   **pH**: Soil pH
-   **SOC**: Soil organic carbon content in the fine earth fraction, g kg^-1^
-   **OCD**: Organic carbon density, kg dm³

We have information per depth in each plot as shown below:

```{r par_soil}
#| message: false
#| warning: false
#| fig-cap: "Soil data variation with depth at the site."
read_tsv("../data/derived_data/soil.tsv") %>%
  filter(site == params$site) %>%
  select(-X, -Y, -ocs, -site) %>%
  gather(variable, value, -depth) %>%
  mutate(var_long = recode(variable,
    "bdod" = '"BD ["~kg~dm^{-3}~"]"',
    "cec" = '"CEC ["~cmol~kg^{-1}~"]"',
    "cfvo" = '"CF ["~cm^3~"100"~cm^{-3}~"]"',
    "clay" = '"Clay ["~"%"~"]"',
    "nitrogen" = '"N ["~g~kg^{-1}~"]"',
    "ocd" = '"OCD ["~kg~dm^{3}~"]"',
    "phh2o" = "pH",
    "sand" = '"Sand ["~"%"~"]"',
    "silt" = '"Silt ["~"%"~"]"',
    "soc" = '"SOC ["~g~kg^{-1}~"]"'
  )) %>%
  ggplot(aes(depth, value)) +
  geom_line() +
  facet_wrap(~var_long, scales = "free_y", labeller = label_parsed) +
  theme_bw() +
  ylab("") +
  xlab("Depth [ m ]") +
  scale_color_discrete(guide = "none") +
  theme(strip.text = element_text(size = 7))
```

We summarised the **two top horizons (0-15cm)** revealing most inter-site variations (V. Freycon pers. com.):

```{r allplot5}
#| message: false
#| warning: false
#| fig-cap: "Mean soil data per site for top horizon (up to 5cm)."
read_tsv("../data/derived_data/soil.tsv") %>%
  filter(site == params$site) %>%
  select(-X, -Y, -site) %>%
  filter(depth <= 15) %>%
  select(-depth) %>%
  summarise_all(mean, na.omit = TRUE) %>%
  gather(variable, value) %>%
  mutate(acronym = recode(variable,
    "bdod" = "BD",
    "cec" = "CEC",
    "cfvo" = "CF",
    "clay" = "Clay",
    "nitrogen" = "N",
    "ocd" = "OCD",
    "ocs" = "OCS",
    "phh2o" = "pH",
    "sand" = "Sand",
    "silt" = "Silt",
    "soc" = "SOC"
  )) %>%
  mutate(unit = recode(variable,
    "bdod" = "$kg~dm^{-3}$",
    "cec" = "$cmol~kg^{-1}$",
    "cfvo" = "$cm^3~100~cm^{-3}$",
    "clay" = "%",
    "nitrogen" = "$g~kg^{-1}$",
    "ocd" = "$kg~dm^{3}$",
    "ocs" = "$kg~m^{-2}$",
    "phh2o" = "",
    "sand" = "%",
    "silt" = "%",
    "soc" = "$g~kg^{-1}$"
  )) %>%
  mutate(variable = recode(variable,
    "bdod" = "Bulk density of the fine earth fraction",
    "cec" = "Cation Exchange Capacity of the soil",
    "cfvo" = "Volumetric fraction of coarse fragments",
    "clay" = "Proportion of clay particles",
    "nitrogen" = "Total nitrogen",
    "ocd" = "Organic carbon density",
    "ocs" = "Organic carbon stocks",
    "phh2o" = "Soil pH",
    "sand" = "Proportion of sand particles",
    "silt" = "Proportion of silt particles",
    "soc" = "Soil organic carbon content in the fine earth fraction"
  )) %>%
  select(acronym, variable, value, unit) %>%
  arrange(acronym) %>%
  kable(
    caption = "Mean soil data of the site for top horizon (up to 15cm).",
    digits = 2
  )
```

This can be represented as a soil texture plot:

```{r betterplot}
#| message: false
#| warning: false
#| fig-cap: "Soil texture for all depths."
read_tsv("../data/derived_data/soil.tsv") %>%
  filter(site == params$site) %>%
  ggplot(aes(x = sand, y = clay, z = silt, col = depth)) +
  ggtern::coord_tern(L = "x", T = "y", R = "z") +
  geom_point() +
  theme_bw() +
  labs(
    yarrow = "Clay [ % ]",
    zarrow = "Silt [ % ]",
    xarrow = "Sand [ % ]"
  ) +
  ggtern::theme_showarrows() +
  ggtern::theme_hidetitles() +
  ggtern::theme_clockwise() +
  scale_color_viridis_c("Depth [ m ]")
```

## References
